soa

### 2024-09-24 inst/manuscript_figures.Rmd
resolved conflicts

# merge_ds <- function(a, b) {
#   # merge sample tables
#   sample_cols <- intersect(colnames(a$samples), colnames(b$samples))
#   s <- bind_rows(select(a$samples, all_of(sample_cols)),
#                  select(b$samples, all_of(sample_cols)))
#   
#   # merge count data
#   a_counts <- bind_cols(select(a$meta, probe, gene, bg), a$counts)
#   b_counts <- bind_cols(select(b$meta, probe, gene, bg), b$counts)
#   # Madison edit: added suffix to avoid addition of ".y" to the end of column names
#   counts <- inner_join(a_counts, b_counts, by=c("probe", "gene", "bg"), suffix = c("", ""),
#                        relationship="one-to-one")
# 
#   # break into metadata and count data
#   meta <- select(counts, -all_of(s$aoi))
#   counts <- select(counts, all_of(s$aoi))
# 
#   # make new dataset object
#   ds <- stgeomx::init(samples=s, meta=meta, counts=counts, x=counts)
#   return(ds)
# }

@@

# merge datasets (all annotated data)
ds <- stgeomx::merge_ds(ds, ds_ln)
ds <- stgeomx::merge_ds(ds, ds_colon)
ds <- stgeomx::merge_ds(ds, ds_brain)
ds <- stgeomx::merge_ds(ds, ds_kidney)

@@

## QC Figures: AOI filtering, results stratified by different parameters
```{r qc filter aoi, stratification}
ds_func <- ds_kidney

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect2(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

pca <- function(s, y) {
  # PCA
  pca_res <- prcomp(t(y), center=TRUE, scale.=FALSE)
  pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
  pca_res$var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

@@

## FIGURE 2: effects of bg correction and normalization on merged only
```{r bg correction and normalization methods plot}
correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

@@

# plot background correction
plot_bgsub_qn <- plot_bgcorrect_helper(fds_bgsub_qn)
plot_bgsub_rle <- plot_bgcorrect_helper(fds_bgsub_rle)
plot_bgsub_q3 <- plot_bgcorrect_helper(fds_bgsub_q3)
plot_bgsub_tmm <- plot_bgcorrect_helper(fds_bgsub_tmm)

plot_qq_qn <- plot_bgcorrect_helper(fds_qq_qn)
plot_qq_rle <- plot_bgcorrect_helper(fds_qq_rle)
plot_qq_q3 <- plot_bgcorrect_helper(fds_qq_q3)
plot_qq_tmm <- plot_bgcorrect_helper(fds_qq_tmm)

plot_none_qn <- plot_bgcorrect_helper(fds_none_qn)
plot_none_rle <- plot_bgcorrect_helper(fds_none_rle)
plot_none_q3 <- plot_bgcorrect_helper(fds_none_q3)
plot_none_tmm <- plot_bgcorrect_helper(fds_none_tmm)

@@

f <- file.path(correction_normalization_dir, "bgcorrect_normalization_grid")
ggsave(paste0(f, ".png"), p, width=20, height=15)
ggsave(paste0(f, ".pdf"), p, width=20, height=15)
```

@@

### Chi-squared analysis of clusters
``` {r}
# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
chi_squared_stats <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")

# Loop through each method
for (i in seq_along(methods)) {
  # Create the contingency table
  contingency_table <- table(methods[[i]]$samples$study, methods[[i]]$samples$cluster_name)

@@

@@

# Format p-values to scientific notation
p_values_formatted <- format(p_values, scientific = TRUE)

@@

@@

```{r}
de_setup <- function(method_name, de_padj_cutoff=0.05, de_log2fc_cutoff=1) {
  # access fds and log2_norm_cpm
  fds_name <- paste0("fds_", method_name)
  fds <- get(fds_name)
  log2_name <- paste0("log2_norm_cpm_", method_name)
  log2_norm_cpm <- get(log2_name)

@@

  # write cluster results
  file_name <- paste0("de_clusters_", method_name, ".xlsx")
  write_xlsx(res, file.path(results_dir, file_name))
  return(res)
}

@@

# res_none_q3 <- de_setup("none_q3")
# res_none_tmm <- de_setup("none_tmm")
# res_none_rle <- de_setup("none_rle")
# res_none_qn <- de_setup("none_qn")
# 
# res_bgsub_q3 <- de_setup("bgsub_q3")
# res_bgsub_tmm <- de_setup("bgsub_tmm")
# res_bgsub_rle <- de_setup("bgsub_rle")
# res_bgsub_qn <- de_setup("bgsub_qn")
# 
# res_qq_q3 <- de_setup("qq_q3")
# res_qq_tmm <- de_setup("qq_tmm")
# res_qq_rle <- de_setup("qq_rle")
# res_qq_qn <- de_setup("qq_qn")

# write tables
# table(res_none_q3$contrast, res_none_q3$de)

@@

de <- NULL
de <- bind_rows(de, res_qq_qn)

@@

## FIGURE 6: GSEA
``` {r gsea}
# read msigdb data
f <- file.path(data_dir, "msigdb.v2023.2.Hs.json")
msigdb_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
msigdb_genes <- intersect(fds$meta$gene, msigdb_gene_sets$gene_symbol)
# filter genes not in our dataset
msigdb_gene_sets <- msigdb_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)

# gene set categories
gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)

# pass log fold change into gsea with hallmark gene sets (sending new code)

get_de_ranks <- function(xde, mycontrast) {
  ranks <- xde %>% 
    dplyr::filter(contrast == mycontrast) %>%
    select(gene, log2fc, padj) %>%
    mutate(rank = log2fc)
  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
  return(ranks)
}

plot_gsea_enrichment <- function(x, ranks, a, gs) {
  txt_stat <- paste0("ES=", round(x$ES,2), " NES=", round(x$NES, 2), " padj=", format(x$padj, scientific=TRUE, digits=3))
  txt_title <- paste0("ranks: ", a, " gs: ", x$pathway)
  p <- plotEnrichment(gs[[x$pathway]], ranks) +
    annotate(geom="text", x=150, y=0.1, label=txt_stat, hjust=0) +
    labs(title = txt_title, 
         xlab = "Rank",
         ylab = "Enrichment Score") +
    theme(plot.title = element_text(size=8))
  return(p)
}

# fgsea_run_batch <- function(xde, xgs, 
#                             fgsea_eps=0,
#                             fgsea_nperms=10000, 
#                             fgsea_nproc=0) {
#   fgsea_scoretypes <- c("pos", "neg")
#   gsea <- NULL
#   for (xcontrast in unique(xde$contrast)) {
#     ranks <- get_de_ranks(xde, xcontrast)
#     for (fgsea_scoretype in fgsea_scoretypes) {
      res <- fgsea(pathways=xgs, 
                   stats=ranks, 
                   eps=fgsea_eps, 
                   scoreType = fgsea_scoretype,
                   nPermSimple=fgsea_nperms, 
                   nproc=fgsea_nproc)
      res <- mutate(res, 
                    contrast = xcontrast, 
                    dir = fgsea_scoretype)
      gsea <- bind_rows(gsea, res)      
#     }
#   }
#   return(gsea)
# }

#
# Run full analysis
#
gsea <- NULL

# hallmark
res <- fgsea_run_batch(de, gs_hallmark, 
                       fgsea_eps=0, fgsea_nperms=gsea_nperms, fgsea_nproc=10)
gsea <- bind_rows(gsea, res)
gsea_all <- gsea
```

# Downstream data analysis 
## Highly variable genes: updated
```{r updated highly variable genes, include=FALSE}
most_var_genes <- function(x, ntop=500, span=0.50) {
  tbl <- bind_cols(
    u = rowMeans(x),
    v = apply(x, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=span)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v - tbl$lovar
  keep <- order(tbl$vadj, decreasing=TRUE)[seq_len(min(ntop, nrow(tbl)))]
  return(x[keep,])
}

maplot <- function(y, ntop=1000, nlabel=100, myspan=0.5) {
  tbl <- bind_cols(
    g = rownames(y),
    u = rowMeans(y),
    v = apply(y, 1, function(x) var(x))
  )
  lovar <- loess(v ~ u, data=tbl, span=myspan)
  tbl$lovar <- predict(lovar)
  tbl$vadj <- tbl$v - tbl$lovar
  tbl <- tbl %>% 
    mutate(rank = min_rank(-vadj)) %>%
    arrange(rank)
  label_data <- filter(tbl, rank <= nlabel)
  top_data <- filter(tbl, rank < ntop)
  bot_data <- filter(tbl, rank >= ntop)
  p <- ggplot(tbl) +
    geom_point(data=bot_data, aes(u, v), alpha=0.5, size=1, color="#aaaaaa") +
    geom_point(data=top_data, aes(u, v), alpha=0.8, size=2, color="#ffaaaa") +
    geom_line(aes(u, lovar), color="blue") +
    geom_text_repel(data=label_data, aes(u, v, label=g), color="black", size=3, max.overlaps=Inf) +
    theme_minimal()
  return(p)
}

heatmap_genes <- function(s, x, filename="heatmap.pdf", width=10, height=10) {
  s <- s %>% mutate(
    noise_quantile = ntile(desc(bg_auc), 8)
  )
  annot_col <- column_to_rownames(s, "aoi") %>%
    select(patient, path_specimen, path_aoi, noise_quantile) %>%
    as.data.frame()
  p <- pheatmap(x,
                scale="row",
                show_rownames=TRUE,
                show_colnames=FALSE,
                border_color=NA,
                clustering_distance_rows = "euclidean",
                clustering_distance_cols = "euclidean",
                clustering_method = "ward.D2",
                # clustering_method = "average",
                # clustering_distance_rows = "correlation",
                # clustering_distance_cols = "correlation",
                breaks = seq(-3, 3, 0.05),
                color = colorRampPalette(c("purple3", "black", "yellow2"))(120),
                annotation_col = annot_col,
                # annotation_colors = color_scales,
                fontsize=4,
                filename=filename,
                width=width,
                height=height)
  return(p)
}

y <- log2_norm_cpm_merged
ntop <- 1000
nlabel <- 100
myspan <- 0.5

# maplot
p <- maplot(log2_norm_cpm_merged, ntop=ntop, nlabel=nlabel, myspan=myspan)
f <- file.path(plot_dir, paste0("maplot"))
ggsave(paste0(f, ".pdf"), plot=p, width=10, height=10)
ggsave(paste0(f, ".png"), plot=p, width=10, height=10)

# select most variable genes
ntop <- 500
myspan <- 0.5
s <- fds$samples
y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)

# heatmap
f <- file.path(plot_dir, "heatmap_highvar_genes.pdf")
p <- heatmap_genes(s, y, f)
```

## PCA: updated
```{r updated pca}
# subset most variable genes
ntop <- 250
myspan <- 0.5
s <- fds$samples
y <- most_var_genes(log2_norm_cpm_merged, ntop=ntop, span=myspan)

# plot dimensions
width <- 12
height <- 10

# PCA
npcs <- 50
pca_res <- prcomp(t(y), center=TRUE, scale.=TRUE, rank.=npcs)
pca_res.var <- round(100 * pca_res$sdev^2 / sum( pca_res$sdev^2 ), 2)
pca_res.var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

# fviz plots
fviz_ntop <- 50

# percent variance of each PC
p <- fviz_eig(pca_res, col.var="blue", addlabels=TRUE)
f <- file.path(plot_dir, paste0("pca_fviz_eig.pdf"))
ggsave(f, p)

# contributions of variables to each PC
p <- fviz_contrib(pca_res, choice="var", axes=1, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib1.pdf"))
ggsave(f, p)

p <- fviz_contrib(pca_res, choice="var", axes=2, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib2.pdf"))
ggsave(f, p)

p <- fviz_contrib(pca_res, choice="var", axes=3, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib3.pdf"))
ggsave(f, p)

# contribution to first 3 PCs
p <- fviz_contrib(pca_res, choice="var", axes = 1:3, top=fviz_ntop) +
  theme(axis.text.x = element_text(angle=90))
f <- file.path(plot_dir, paste0("pca_fviz_contrib123.pdf"))
ggsave(f, p)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(1,2),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc12.pdf"))
ggsave(f, p, width=7, height=7)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(1,3),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc13.pdf"))
ggsave(f, p, width=7, height=7)

# graph of variables contributing to PCA
p <- fviz_pca_var(
  pca_res,
  axes = c(2,3),
  col.var = "contrib", # Color by the quality of representation
  gradient.cols = c("darkorchid4", "gold", "darkorange"),
  repel=TRUE,
  select.var = list(contrib=fviz_ntop)
)
f <- file.path(plot_dir, paste0("pca_fviz_pca_var_pc23.pdf"))
ggsave(f, p, width=7, height=7)

#
# PCA plots
#
x <- bind_cols(s, as_tibble(pca_res$x))

>>>>

@@

# merge datasets (all annotated data)
ds <- stgeomx::merge(ds, ds_ln)
ds <- stgeomx::merge(ds, ds_colon)
ds <- stgeomx::merge(ds, ds_brain)
ds <- stgeomx::merge(ds, ds_kidney)

@@

## QC Figures: AOI filtering, results stratified by different parameters
```{r qc filter aoi, stratification}
ds_func <- ds_brain

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_merged")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_panc)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_panc")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_ln)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_ln")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_kidney)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_kidney")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_colon)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_colon")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

# plot background correction (before and after)
p <- stgeomx::plot_bgcorrect(fds_brain)
p
w <- 8
h <- 8
f <- file.path(correction_normalization_dir, "qc_bgcorrect_brain")
ggsave(paste0(f, ".png"), p, width=w, height=h)
ggsave(paste0(f, ".pdf"), p, width=w, height=h)

@@

pca <- function(s, y) {
  # PCA
  pca_res <- prcomp(t(y), center=TRUE, scale.=FALSE)
  pca_res.var <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2 ), 2)
  pca_res$var_txt <- paste(colnames(pca_res$x), " (", paste(as.character(pca_res.var), "%", ")", sep=""), sep="")

@@

## FIGURE 2: effects of bg correction and normalization on merged only
```{r bg correction and normalization methods plot}
library("patchwork")
correction_normalization_dir <- file.path(plot_dir, "bgcorrection_normalization")

@@

# plot background correction
plot_bgsub_qn <- stgeomx::plot_bgcorrect(fds_bgsub_qn)
plot_bgsub_rle <- stgeomx::plot_bgcorrect(fds_bgsub_rle)
plot_bgsub_q3 <- stgeomx::plot_bgcorrect(fds_bgsub_q3)
plot_bgsub_tmm <- stgeomx::plot_bgcorrect(fds_bgsub_tmm)

plot_qq_qn <- stgeomx::plot_bgcorrect(fds_qq_qn)
plot_qq_rle <- stgeomx::plot_bgcorrect(fds_qq_rle)
plot_qq_q3 <- stgeomx::plot_bgcorrect(fds_qq_q3)
plot_qq_tmm <- stgeomx::plot_bgcorrect(fds_qq_tmm)

plot_none_qn <- stgeomx::plot_bgcorrect(fds_none_qn)
plot_none_rle <- stgeomx::plot_bgcorrect(fds_none_rle)
plot_none_q3 <- stgeomx::plot_bgcorrect(fds_none_q3)
plot_none_tmm <- stgeomx::plot_bgcorrect(fds_none_tmm)

@@

f <- file.path(correction_normalization_dir, "bgcorrect_normalization_grid")
ggsave(paste0(f, ".png"), p, width=30, height=15)
ggsave(paste0(f, ".pdf"), p, width=30, height=15)
```

@@

### Cluster purity
```{r purity}
library("IntNMF")

# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
purity_table <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
purity_table <- NULL

# Loop through each method
for (i in methods) {

  # create purity score and add to purity table
  purity_table <- rbind(purity_table, ClusterPurity(i$samples$study, i$samples$cluster_name))
}

# Combine results into a data frame
results_df <- data.frame(
  Method = method_names,
  Purity_Scores = purity_table
 )

### Chi-squared analysis of clusters
``` {r}
library(reshape2)

# List of methods
methods <- list(fds_none_q3, fds_none_tmm, fds_none_rle, fds_none_qn, 
                fds_bgsub_q3, fds_bgsub_tmm, fds_bgsub_rle, fds_bgsub_qn, 
                fds_qq_q3, fds_qq_tmm, fds_qq_rle, fds_qq_qn)

# Initialize vectors to store Chi-squared statistics and p-values
chi_squared_stats <- numeric(length(methods))
method_names <- c("none_q3", "none_tmm", "none_rle", "none_qn",
                   "bgsub_q3", "bgsub_tmm", "bgsub_rle", "bgsub_qn",
                   "qq_q3", "qq_tmm", "qq_rle", "qq_qn")
master_contingency_table <- NULL

# Loop through each method
for (i in seq_along(methods)) {
  # Create the contingency table
  contingency_table <- table(methods[[i]]$samples$study, methods[[i]]$samples$cluster_name)
  colnames(contingency_table) <- paste0(colnames(contingency_table), "_", method_names[i])

@@

  # add to master contingency table
  if (i == 1) {
    master_contingency_table <- contingency_table
  }
  else {
    master_contingency_table <- cbind(master_contingency_table, contingency_table)
  }

@@

@@

print(master_contingency_table)
# heatmap(master_contingency_table)

# # Dummy data
# x <- LETTERS[1:20]
# y <- paste0("var", seq(1,20))
# data <- expand.grid(X=x, Y=y)
# data$Z <- runif(400, 0, 5)

# # Heatmap 
# master_contingency_table <- melt(master_contingency_table)
# ggplot(master_contingency_table, aes(x = Var1, y = Var2, fill = value)) +
# geom_tile()

@@

```{r}
library("limma")
library("writexl")

de_setup <- function(method_name, de_padj_cutoff=0.05, de_log2fc_cutoff=1) {
  # access fds and log2_norm_cpm
  fds_name <- paste0("fds_", method_name)
  fds <- get(fds_name)
  log2_name <- paste0("log2_norm_cpm_", method_name)
  log2_norm_cpm <- get(log2_name)

@@

  # write cluster results
  file_name <- paste0("de_clusters_", method_name, ".xlsx")
  write_xlsx(res, file.path(results_dir, file_name))
  de <- bind_rows(de, res)
  return(de)
}

@@

de_none_q3 <- de_setup("none_q3")
de_none_tmm <- de_setup("none_tmm")
de_none_rle <- de_setup("none_rle")
de_none_qn <- de_setup("none_qn")

de_bgsub_q3 <- de_setup("bgsub_q3")
de_bgsub_tmm <- de_setup("bgsub_tmm")
de_bgsub_rle <- de_setup("bgsub_rle")
de_bgsub_qn <- de_setup("bgsub_qn")

de_qq_q3 <- de_setup("qq_q3")
de_qq_tmm <- de_setup("qq_tmm")
de_qq_rle <- de_setup("qq_rle")
de_qq_qn <- de_setup("qq_qn")

# write tables
table(de_none_q3$contrast, de_none_q3$de)

@@

de <- NULL
de <- de_qq_qn

@@

## FIGURE 6: GSEA
``` {r gsea}
gsea_nperms <- 10000

read_msigdb_json_file <- function(filename) {
  x <- jsonlite::fromJSON(filename)
  y <- tibble(gs_name = names(x), gs = x) %>% 
    unnest_wider(gs) %>%
    unnest_longer(geneSymbols) %>%
    select(gs_cat = collection,
           gs_name = gs_name,
           gene_symbol = geneSymbols)
}

# read msigdb data: hallmark
f <- system.file("extdata", "h.all.v2024.1.Hs.json", package = "stgeomx")
msigdb_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
msigdb_genes <- intersect(fds_qq_qn$meta$gene, msigdb_gene_sets$gene_symbol)
# filter genes not in our dataset
msigdb_gene_sets <- msigdb_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_hallmark <- dplyr::filter(msigdb_gene_sets, gs_cat == "H")
gs_hallmark <- split(x = gs_hallmark$gene_symbol, f = gs_hallmark$gs_name)

# read msigdb data: pancreas_endocrine
f <- system.file("extdata", "GOBP_ENDOCRINE_PANCREAS_DEVELOPMENT.v2024.1.Hs.json", package = "stgeomx")
gobp_end_pancreas_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_end_pancreas_genes <- intersect(fds_qq_qn$meta$gene, gobp_end_pancreas_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_end_pancreas_gene_sets <- gobp_end_pancreas_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_end_panc <- dplyr::filter(gobp_end_pancreas_gene_sets, gs_cat == "C5:GO:BP")
gs_end_panc <- split(x = gs_end_panc$gene_symbol, f = gs_end_panc$gs_name)

# read msigdb data: pancreas secretion
f <- system.file("extdata", "GOBP_INSULIN_SECRETION.v2024.1.Hs.json", package = "stgeomx")
gobp_pancreas_secretion_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_pancreas_secretion_genes <- intersect(fds_qq_qn$meta$gene, gobp_pancreas_secretion_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_pancreas_secretion_gene_sets <- gobp_pancreas_secretion_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_panc_sec <- dplyr::filter(gobp_pancreas_secretion_gene_sets, gs_cat == "C5:GO:BP")
gs_panc_sec <- split(x = gs_panc_sec$gene_symbol, f = gs_panc_sec$gs_name)

# read msigdb data: brain development gene sets
f <- system.file("extdata", "brain_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_brain_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_brain_genes <- intersect(fds_qq_qn$meta$gene, gobp_brain_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_brain_gene_sets <- gobp_brain_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_brain <- dplyr::filter(gobp_brain_gene_sets, gs_cat == "C5:GO:BP")
gs_brain <- split(x = gs_brain$gene_symbol, f = gs_brain$gs_name)

# read msigdb data: ln development gene sets
f <- system.file("extdata", "ln_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_ln_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_ln_genes <- intersect(fds_qq_qn$meta$gene, gobp_ln_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_ln_gene_sets <- gobp_ln_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_ln <- dplyr::filter(gobp_ln_gene_sets, gs_cat == "C5:GO:BP")
gs_ln <- split(x = gs_ln$gene_symbol, f = gs_ln$gs_name)

# read msigdb data: kidney development gene sets
f <- system.file("extdata", "kidney_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_kidney_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_kidney_genes <- intersect(fds_qq_qn$meta$gene, gobp_kidney_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_kidney_gene_sets <- gobp_kidney_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_kidney <- dplyr::filter(gobp_kidney_gene_sets, gs_cat == "C5:GO:BP")
gs_kidney <- split(x = gs_kidney$gene_symbol, f = gs_kidney$gs_name)

# read msigdb data: colon development gene sets
f <- system.file("extdata", "colon_genesets.v2024.1.Hs.json", package = "stgeomx")
gobp_colon_gene_sets <- read_msigdb_json_file(f)
# determine gene universe
gobp_colon_genes <- intersect(fds_qq_qn$meta$gene, gobp_colon_gene_sets$gene_symbol)
# filter genes not in our dataset
gobp_colon_gene_sets <- gobp_colon_gene_sets %>%
  dplyr::filter(gene_symbol %in% fds_qq_qn$meta$gene) %>%
  distinct(gs_cat, gs_name, gene_symbol, .keep_all = TRUE)
# gene set categories
gs_colon <- dplyr::filter(gobp_colon_gene_sets, gs_cat == "C5:GO:BP")
gs_colon <- split(x = gs_colon$gene_symbol, f = gs_colon$gs_name)

# new gsea code
get_fgsea_ranks_de <- function(xde, mycontrast) {
  ranks <- xde %>%
    dplyr::filter(contrast == mycontrast) %>%
    select(gene, log2fc, padj) %>%
    mutate(rank = log2fc)
  ranks = sort(setNames(ranks$rank, ranks$gene), decreasing = TRUE)
  return(ranks)
}

fgsea_run_batch <- function(xde, xgs,
                            fgsea_eps=0,
                            fgsea_nperms=1000) {
  gsea <- NULL
  for (xcontrast in unique(xde$contrast)) {
    ranks <- get_fgsea_ranks_de(xde, xcontrast)
    res <- fgsea(pathways=xgs,
                 stats=ranks,
                 eps=fgsea_eps,
                 nPermSimple=fgsea_nperms)
    res <- mutate(res, contrast = xcontrast)
    gsea <- bind_rows(gsea, res)     
  }
  return(gsea)
}

stgeomx_gsea_pipeline <- function(gsea, de, method) {
  res <- fgsea_run_batch(de, gs_hallmark,
                       fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_ex_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_end_panc,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_panc_sec,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_brain,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_ln,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_kidney,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)

  res <- fgsea_run_batch(de, gs_colon,
                         fgsea_eps=0, fgsea_nperms=gsea_nperms)
  res <- res %>% mutate(method = method)
  gsea <- bind_rows(gsea, res)
